Synthesizing Quotient Lenses

### 2018-03-09 papers/QuotientLenses/main.tex
Checkpoint

QRE lenses address some of the limitations of bijective lenses because a single
lens program expresses both the canonizers and the transformation between kernel
languages simultaneously, which reduces programmer effort.  But we can go even
further by recognizing that the type structure of QRE lenses contains
information that can be exploited to automatically synthesize lenses from their
types.  Rather than writing the QRE lens manually, the programmer can instead
specify the desired behavior of a lens by giving its interface types and providing
examples, if necessary, to disambiguate among possible implementations.  This
way of constructing lenses can often be much simpler than building them by
hand---see the example of lens synthesis in
Section~\ref{sec:examplesynth}.

The Optician framework~\cite{optician} showed how to do such lens synthesis in
the case for bijective lenses. Here we show how to reduce QRE lens synthesis to
that case, so that we can re-use the Optician algorithm but in the more
expressive context of QRE lenses.  The basic idea is straightforward: we run the
Optician algorithm to synthesize a lens between the kernels of two QREs and then
apply the canonizers at the edges to recover a lens between the whole
languages.  This simple strategy turns out to be remarkably effective, and the
idea of using Optician in this way is part the impetus for the design of our QRE
lenses.

The Optician synthesis algorithm works as follows. Given regular expressions $R$
and $S$ and a set of example input--output pairs
$\{(r_1, s_1),\allowbreak \ldots,\allowbreak (r_n, s_n)\}$, Optician will try to
find a bijective lens $\ell : R \Leftrightarrow S$ that that agrees on the
examples, \textit{i.e.}, it maps $r_i$ to $s_i$ in the forward direction and
$s_i$ to $r_i$ in the backward direction.  The Optician algorithm is guaranteed
to succeed if such a lens exists.    \saz{We should probably mention what
happens if the examples aren't sufficient to uniquely determine the lens.}

In our setting, we want to synthesize a quotient lens $q: c_1 \Leftrightarrow
c_2$ from the QREs $c_1$ and $c_2$ and a set of example input-output pairs
$\{(x_1, y_1),\allowbreak \ldots,\allowbreak (x_n, y_n)\}$ where the $x_i$'s are
in $W(c_1)$ and the $y_i$'s are in $W(c_2)$. We furthermore wish $q$ to map the
equivalence class of $x_i$ to the equivalence class of $y_i$ and vice versa:
\begin{align*}
q.\get(x_i) &\equiv_{c_2} y_i \text{, and}\\
q.\lput(y_i) &\equiv_{c_1} x_i
\end{align*}
Our approach to synthesizing QRE lenses is guided by Theorem~\ref{normal form},
which says that, if there is a derivation $q : c_1 \Leftrightarrow c_2$ of a
QRE lens, then there exists a bijective lens $\ell : K(c_1) \Leftrightarrow
K(c_2)$ such that
\begin{align*}
\llbracket q \rrbracket.\get &= \llbracket \ell \rrbracket\circ \canonizer(c_1)\\
\llbracket q \rrbracket.\lput &= \llbracket \ell \rrbracket^{-1} \circ
\canonizer(c_2).
\end{align*}

For the QRE examples, the $x_i$'s are in $W(c_1)$ and the $y_i$'s are
in $W(c_2)$, so we can construct ${x'_i} = \canonizer(c_1)(x_i)$ in $K(c_1)$
and ${y'_i} = \canonizer(c_2)(y_i)$ is in $K(c_2)$. To synthesize the
desired quotient lens $q: c_1 \Leftrightarrow c_2$ that is consistent with the
input-output examples $\{(x_1, y_1), \ldots, (x_n, y_n)\}$ it suffices to
synthesize a bijective lens $\ell : K(c_1) \Leftrightarrow K(c_2)$ that is
consistent with the canonized examples $\{({x'_1}, {y'_1}), \ldots, ({x'_n},
{y'_n})\}$ and then apply the canonizers at the outside. Our procedure for QRE
lens synthesis is thus
\begin{enumerate}
  \item
  Compute $K(c_1)$, $K(c_2)$,
  \item
  Compute $\{({x'_1}, {y'_1}), \ldots, ({x'_n}, {y'_n})\}$ where ${x'_i} =
  \canonizer(c_1)(x_i)$ and ${y'_i} = \canonizer(c_2)(y_i)$,
\item Synthesize, using Optician's algorithm, a bijective lens
  $\ell : K(c_1) \Leftrightarrow K(c_2)$ that is consistent with the examples
  $\{({x'_1}, {y'_1}),\allowbreak \ldots,\allowbreak ({x'_n}, {y'_n})\}$
  \item
  Return $q = \kw{rquot}(\kw{lquot}(c_1, \ell), c_2) : c_1 \Leftrightarrow
  c_2$
\end{enumerate}

@@

\begin{minipage}{3in}
\begin{lstlisting}
bib_example =
"@Book {Doe,
 Author = \"John Doe\",
 Title = {Generic Title},
}"
\end{lstlisting}
\end{minipage}
\begin{minipage}{3in}
\begin{lstlisting}
end_example =
"%0 Book
 %T Generic Title
 %A John Doe
 %F Doe"
\end{lstlisting}
\end{minipage}

>>>>

QRE lenses address some of the limitations of bijective lenses because a single
lens program expresses both the canonizers and the transformation between kernel
languages simultaneously, which reduces programmer effort.  But we can go even
further by recognizing that the type structure of QRE lenses contains
information that can be exploited to automatically synthesize lenses from their
types.  Rather than writing the QRE lens manually, the programmer can instead
specify the desired behavior of a lens by giving its interface types and providing
examples, if necessary, to disambiguate among possible implementations.  This
way of constructing lenses can often be much simpler than building them by
hand, as we saw in Section~\ref{sec:examplesynth}.

The Optician framework~\cite{optician} showed how to do such lens synthesis in
the case for bijective lenses. Here we show how to reduce QRE lens synthesis to
that case, so that we can re-use the Optician algorithm but in the more
expressive context of QRE lenses.  The basic idea is straightforward: we run the
Optician algorithm to synthesize a lens between the kernels of two QREs and then
apply the canonizers at the edges to recover a lens between the whole
languages.  This simple strategy turns out to be remarkably effective, and the
idea of using Optician in this way is part of the impetus\bcp{What was the
  other part??} for the design of our QRE lenses.

The Optician synthesis algorithm works as follows. Given regular expressions $R$
and $S$ and a set of example input--output pairs
$\{(r_1, s_1),\allowbreak \ldots,\allowbreak (r_n, s_n)\}$, Optician will try to
find a bijective lens $\ell : R \Leftrightarrow S$ that that agrees on the
examples, \textit{i.e.}, it maps $r_i$ to $s_i$ in the forward direction and
$s_i$ to $r_i$ in the backward direction.  The Optician algorithm is guaranteed
to succeed (eventually!) if such a lens exists.    \saz{We should probably
  mention what happens if the examples aren't sufficient to uniquely
  determine the lens.}

In our setting, we want to synthesize a quotient lens $q: c_1 \Leftrightarrow
c_2$ from the QREs $c_1$ and $c_2$ and a set of example input-output pairs
$\{(x_1, y_1),\allowbreak \ldots,\allowbreak (x_n, y_n)\}$ where the $x_i$'s are
in $W(c_1)$ and the $y_i$'s are in $W(c_2)$. We furthermore wish $q$ to map the
equivalence class of $x_i$ to the equivalence class of $y_i$ and vice versa:
\begin{align*}
q.\get(x_i) &\equiv_{c_2} y_i \text{, and}\\
q.\lput(y_i) &\equiv_{c_1} x_i
\end{align*}
Our approach to synthesizing QRE lenses is guided by Theorem~\ref{normal form},
which says that, if there is a derivation $q : c_1 \Leftrightarrow c_2$ of a
QRE lens, then there exists a bijective lens $\ell : K(c_1) \Leftrightarrow
K(c_2)$ such that
\begin{align*}
\llbracket q \rrbracket.\get &= \llbracket \ell \rrbracket\circ \canonizer(c_1)\\
\llbracket q \rrbracket.\lput &= \llbracket \ell \rrbracket^{-1} \circ
\canonizer(c_2).
\end{align*}
\bcp{This math display and the previous one are punctuated perfedtly: In the
first, the colon replaces a period at the end of the sentence (so no need
for another period at the end of the display).  In the second, the display
is punctuated as if it were inline.  Please go through and make sure all the
other displayed material follows one or the other of these rules.}

For the\bcp{why ``the''?  Are we talking about some set of QRE examples
  already?} QRE examples, the $x_i$'s are in $W(c_1)$ and the $y_i$'s are 
in $W(c_2)$, so we can construct ${x'_i} = \canonizer(c_1)(x_i)$ in $K(c_1)$
and ${y'_i} = \canonizer(c_2)(y_i)$ is in $K(c_2)$. To synthesize the
desired quotient lens $q: c_1 \Leftrightarrow c_2$ that is consistent with the
input-output examples $\{(x_1, y_1), \ldots, (x_n, y_n)\}$ it suffices to
synthesize a bijective lens $\ell : K(c_1) \Leftrightarrow K(c_2)$ that is
consistent with the canonized examples $\{({x'_1}, {y'_1}), \ldots, ({x'_n},
{y'_n})\}$ and then apply the canonizers at the outside. Our procedure for QRE
lens synthesis is thus
\begin{enumerate}
  \item
  Compute $K(c_1)$, $K(c_2)$,
  \item
  Compute $\{({x'_1}, {y'_1}), \ldots, ({x'_n}, {y'_n})\}$ where ${x'_i} =
  \canonizer(c_1)(x_i)$ and ${y'_i} = \canonizer(c_2)(y_i)$,
\item Synthesize, using Optician's algorithm, a bijective lens
  $\ell : K(c_1) \Leftrightarrow K(c_2)$ that is consistent with the examples
  $\{({x'_1}, {y'_1}),\allowbreak \ldots,\allowbreak ({x'_n}, {y'_n})\}$
  \item
  Return $q = \kw{rquot}(\kw{lquot}(c_1, \ell), c_2) : c_1 \Leftrightarrow
  c_2$ \bcp{punctuate this enumeration as if it were running text: ``, and''
  at the end of (3) and period here.}
\end{enumerate}

@@

\begin{center}
\begin{minipage}{2.5in}
\begin{lstlisting}
bib_example =
"@Book {Doe,
 Author = \"John Doe\",
 Title = {Generic Title},
}"
\end{lstlisting}
\end{minipage}
\begin{minipage}{1.5in}
\begin{lstlisting}
end_example =
"%0 Book
 %T Generic Title
 %A John Doe
 %F Doe"
\end{lstlisting}
\end{minipage}
\end{center}
